Predicting Bike Usage at UofT St. George with Bayesian Poisson Regression

### 2024-11-29 paper/paper.qmd
paper structure is added
paper structure is added

---
title: "Predicting Bike Usage at UofT St. George with Bayesian Poisson Regression"
subtitle: "Usage grows by one every 20 days, drops to nearly zero in winter, with 79% occurring between 8 AM and 8 PM"

author: 
  - Haowei Fan
thanks: "Code and data are available at: https://github.com/HaoweiFan0912/Bikeshare-Forecast.git"
date: today
date-format: long
abstract: "University of Toronto's St. George campus students and staff frequently encounter challenges in locating available bike-sharing parking spots upon arrival and bikes upon leaving [@elassi2017bike], underscoring the necessity for accurate predictions of future bike-sharing demand to optimize campus commuting infrastructure. This study employs Bayesian Poisson regression to predict the utilization of bike-sharing stations at 27 locations across campus for specific years, months, and four-hour intervals of the day. The results indicate that bike-sharing usage across campus is experiencing rapid growth, with a sharp decline during the winter months and 79.8% of peak usage consistently observed between 8:00 AM and 8:00 PM daily. Specifically, on September 26, 2025, between 4:00 and 8:00 AM, the usage of 8 stations is projected to increase by 3 bikes, 14 stations by 2 bikes, 3 stations by 1 bike, while 1 station will remain unchanged, and 1 station will see a decrease by 1 bike, compared to the same timeframe in 2024."
format: pdf
toc-depth: 2    
number-sections: true
toc: true           
toc-title: "Table of Contents" 
bibliography: references.bib
---

@@

## Overview
Bike-sharing services have experienced rapid growth in the Toronto region, with ridership rising dramatically from approximately 665,000 trips in 2015 to over 4.5 million in 2022, according to a study by the University of Toronto's School of Cities[@liu2023bike]. This rapid increase highlights the critical need for data-driven analyses to optimize bike-sharing systems and guide policy decisions that can further enhance urban mobility[@xiamen]. While city-wide studies provide valuable insights, research from Hangzhou, China, has demonstrated that bike-sharing usage patterns can vary significantly between urban areas and university campuses[@hangzhou]. These differences suggest that general urban studies conducted in the Toronto region may not fully capture the unique dynamics of campus-based bike-sharing systems. Therefore, to address this gap in the literature, this study focuses on the University of Toronto St. George Campus as a case study, analyzing bike-sharing usage within the campus in Toronto to provide a nuanced understanding of its role in campus mobility and inform policies tailored to similar settings.

@@

This study found that although the usage at all stations within the University of Toronto campus has increased by an average of one user every 20 days each year, there are significant seasonal differences. During winter, usage at all stations, regardless of the year, remains close to zero. Additionally, from January to September each year, the usage shows an upward trend, which then gradually declines over the following months. There are also significant differences in usage across different times of the day. The period from 8 am to 8 pm accounts for an average of 79.8% of daily usage, while the period from midnight to 8 am accounts for only 6.8%. It is noteworthy that this project forecasts an average increase of 2 uses across 27 stations between 4:00 AM and 8:00 AM on September 26, 2025, compared to the same time period in 2024. Specifically, usage at 8 stations is expected to increase by 3, 14 stations by 2, and 3 stations by 1, while 1 station will remain unchanged and another will decrease by 1.

Compared to the general statistics on bike-sharing usage in the Greater Toronto Area, these findings offer more detailed recommendations specifically for transportation planning and policy adjustments within the University of Toronto campus. They also provide guidance on how to better allocate, deploy, or store shared bikes.

** ToBeMoreSpecific**
The paper proceeds as follows: @sec-data outlines the data sources, processing techniques, and the variables employed in the study. @sec-model describes the Bayesian Poisson regression model used for prediction, including its formulation and validation. @sec-result discusses the results, emphasizing temporal and station-specific patterns in bike-sharing usage. @sec-discussion concludes with a discussion of the study’s contributions, limitations, and recommendations for future research.

>>>>

---
title: "Predicting Bike Usage at UofT St. George with Bayesian Poisson Regression"
subtitle: "Daily usage grows by one every 20 days, drops to between 0 and 3 in winter, with 79% occurring between 8 AM and 8 PM"
author: 
  - Haowei Fan
thanks: "Code and data are available at: https://github.com/HaoweiFan0912/Bikeshare-Forecast.git"
date: today
date-format: long
abstract: "University of Toronto's St. George campus students and staff frequently encounter challenges in locating available bike-sharing parking spots upon arrival and bikes upon leaving [@elassi2017bike], underscoring the necessity for accurate predictions of future bike-sharing demand to optimize campus commuting infrastructure. This study employs Bayesian Poisson regression to predict the utilization of bike-sharing stations at 27 locations across campus for specific years, months, and four-hour intervals of the day. The results show that daily bike-sharing usage on campus increases by one every 20 days but decreases to between 0 and 3 in winter months, and 79% of peak usage consistently observed between 8:00 AM and 8:00 PM daily. Specifically, on September 26, 2025, between 4:00 and 8:00 AM, the usage of 8 stations is projected to increase by 3 bikes, 14 stations by 2 bikes, 3 stations by 1 bike, while 1 station will remain unchanged, and 1 station will see a decrease by 1 bike, compared to the same timeframe in 2024."
format: pdf
toc-depth: 2    
number-sections: true
toc: true           
toc-title: "Table of Contents" 
bibliography: references.bib
---

@@

## Overview
Bike-sharing services have experienced rapid growth in the Toronto region, with ridership rising dramatically from approximately 665,000 trips in 2015 to over 4.5 million in 2022, according to a study by the University of Toronto's School of Cities[@liu2023bike]. This rapid increase highlights the critical need for data-driven analyses to optimize bike-sharing systems and guide policy decisions that can further enhance urban mobility[@xiamen]. While city-wide studies provide valuable suggestions, research from Hangzhou, China, has demonstrated that bike-sharing usage patterns can vary significantly between urban areas and university campuses[@hangzhou]. These differences suggest that general urban studies conducted in the Toronto region may not fully capture the unique dynamics of campus-based bike-sharing systems. Therefore, to address this gap in the literature, this study focuses on the University of Toronto St. George Campus as a case study, analyzing bike-sharing usage within the campus in Toronto to provide a nuanced understanding of its role in campus mobility and inform policies tailored to similar settings.

@@

This study found that although the usage at all stations within the University of Toronto campus has increased by an average of one user every 20 days each year, there are significant seasonal differences. During winter, usage at all stations, regardless of the year, remains close to zero between 0 to 3. Additionally, from January to September each year, the usage shows an upward trend, which then gradually declines over the following months. There are also significant differences in usage across different times of the day. The period from 8 am to 8 pm accounts for an average of 79% of daily usage, while the period from midnight to 8 am accounts for only 6%. It is noteworthy that this project forecasts an average increase of 2 uses across 27 stations between 4:00 AM and 8:00 AM on September 26, 2025, compared to the same time period in 2024. Specifically, usage at 8 stations is expected to increase by 3, 14 stations by 2, and 3 stations by 1, while 1 station will remain unchanged and another will decrease by 1.

Compared to the general statistics on bike-sharing usage in the Greater Toronto Area, these findings offer more detailed recommendations specifically for transportation planning and policy adjustments within the University of Toronto St. George campus. They also provide guidance on how to better allocate, deploy, or store shared bikes.

The structure of this paper is as follows: @sec-data outlines the data sources, evaluates the data measurements, introduces key variables along with their visualizations, presents similar datasets, provides a high-level description of the data cleaning process, and details the technical support utilized in this project. @sec-model provides an in-depth explanation of the model types used in this study and the rationale behind the selection. It introduces the target variable, predictors, and all components of the model. This section also includes model validation and discusses an alternative model. @sec-result uses a specific time frame in 2025 as an example to make predictions using the model, comparing the results with 2024 data. @sec-discussion explores the implications of this study for the world, acknowledges its limitations, and offers recommendations for future research.
